Computing solution space properties of combinatorial optimization problems via generic tensor networks

### 2022-06-07 paper/paper.tex
seperate suppl

\documentclass[onefignum, onetabnum]{siamart190516}
%\usepackage{natbib}
%\usepackage[sort]{cite}
\pdfoutput=1
%\usepackage[colorlinks=true,urlcolor=blue,citecolor=blue,linkcolor=blue]{hyperref}
\usepackage[english]{babel}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amssymb}
\usepackage{tabularx}
\usepackage{quoting}
\usepackage{upquote}
\usepackage{subcaption}
\usepackage{multicol}
\usepackage{cancel}
\usepackage[framemethod=TikZ]{mdframed}
\usetikzlibrary{shapes}
\usetikzlibrary{snakes}
\usetikzlibrary{shapes.geometric}
\usepackage{wrapfig}
%\usepackage{caption}
%\usepackage[plain]{algorithmic}
\usepackage[linesnumbered, ruled, vlined, algo2e]{algorithm2e}
\usepackage{algpseudocode}
\usepackage{rotating}
%\usepackage{cite}
\usepackage{booktabs}
%\usepackage{unicode-math}
%\usepackage{algorithm}% http://ctan.org/pkg/algorithm
%\usepackage{algpseudocode}% http://ctan.org/pkg/algpseudocode
\usepackage{xcolor}% http://ctan.org/pkg/xcolor
\makeatletter
\newsavebox{\@brx}
\newcommand{\llangle}[1][]{\savebox{\@brx}{\(\m@th{#1\langle}\)}%
  \mathopen{\copy\@brx\kern-0.5\wd\@brx\usebox{\@brx}}}
\newcommand{\rrangle}[1][]{\savebox{\@brx}{\(\m@th{#1\rangle}\)}%
  \mathclose{\copy\@brx\kern-0.5\wd\@brx\usebox{\@brx}}}
\makeatother
\usepackage{bbm}
\usepackage{jlcode}
\usepackage{graphicx}
\usepackage{amsmath,color}
\usepackage{mathrsfs}
\usepackage{float}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{indentfirst}
\usepackage{txfonts}
%\usepackage[epsilon, tsrm, altpo]{backnaur}

\makeatletter
\def\parsept#1#2#3{%
    \def\nospace##1{\zap@space##1 \@empty}%
    \def\rawparsept(##1,##2){%
        \edef#1{\nospace{##1}}%
        \edef#2{\nospace{##2}}%
    }%
    \expandafter\rawparsept#3%
}
\makeatother
\DeclareMathAlphabet{\mymathbb}{U}{BOONDOX-ds}{m}{n}
\newcommand{\listingcaption}[1]%
{%
\refstepcounter{lstlisting}\hfill%
Listing \thelstlisting: #1\hfill%\hfill%
}%
\newcolumntype{b}{X}
\newcolumntype{s}{>{\hsize=.7\hsize}X}
\usepackage{listings}
\lstset{
    language=Julia,
    basicstyle=\ttfamily\scriptsize,
    numberstyle=\scriptsize,
    % numbers=left,
    backgroundcolor=\color{gray!7},
    %backgroundcolor=\color{white},
    %frame=single,
    xleftmargin=2em,
    tabsize=2,
    rulecolor=\color{black!15},
    %title=\lstname,
    escapeinside={(*}{*)},
    breaklines=true,
    %breakatwhitespace=true,
    %framextopmargin=2pt,
    %framexbottommargin=2pt,
    frame=bt,
    extendedchars=true,
    inputencoding=utf8,
    columns=fullflexible,
    %escapeinside={(*@}{@*)},
}

\tolerance=1
\emergencystretch=\maxdimen
\hyphenpenalty=1000
\hbadness=1000

\makeatletter

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% User specified LaTeX commands.

%Journal reference.  Comma sets off: name, vol, page, year
\def\journal #1, #2, #3, 1#4#5#6{{\sl #1~}{\bf #2}, #3 (1#4#5#6) }
\def\pr{\journal Phys. Rev., }
\def\prb{\journal Phys. Rev. B, }
\def\prl{\journal Phys. Rev. Lett., }
\def\pl{\journal Phys. Lett., }
%\def\np{\journal Nucl. Phys., }

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%\usepackage[colorlinks, citecolor=blue]{hyperref}
\DeclareMathOperator*{\argmax}{arg\,max}

\newcommand{\eqname}[1]{\stepcounter{equation}\tag{\theequation : #1}}
%%%%%% Shortcut related
\newcommand{\<}{\langle}
\renewcommand{\>}{\rangle}
\newcommand{\out}{{\vx^L}}
\newcommand{\inp}{{\vx^0}}
\newcommand{\cquad}{{{ }_{\quad}}}
\newcommand{\pluseq}{\mathrel{+}=}
\newcommand{\minuseq}{\mathrel{-}=}
\newcommand{\vx}{{\mathbf{x}}}
\newcommand{\vg}{{\mathbf{g}}}
\newcommand{\vp}{{\mathbf{p}}}
\newcommand{\vy}{{\mathbf{y}}}
\newcommand{\Var}{{\mathrm{Var}}}
\newcommand{\Mean}{{\mathrm{E}}}
\newcommand{\vvalue}{{\texttt{value}}}
\newcommand{\grad}{{\texttt{grad}}}
\newcommand{\parameter}{{\texttt{parameter}}}
%%%%%% Convention related
\newcommand{\SWAP}{{\rm SWAP}}
\newcommand{\CNOT}{{\rm CNOT}}
\newcommand{\bigO}{{\mathcal{O}}}
\newcommand{\X}{{\rm X}}
\renewcommand{\H}{{\rm H}}
\newcommand{\Rx}{{\rm Rx}}
\renewcommand{\v}[1]{{\bf #1}}
\newcommand{\dataset}{{\mathcal{D}}}
\newcommand{\wfunc}{{\psi}}
\newcommand{\SU}{{\rm SU}}
\newcommand{\UU}{{\rm U}}
\newcommand{\thetav}{{\boldsymbol{\theta}}}
\newcommand{\gammav}{{\boldsymbol{\gamma}}}
\newcommand{\thetai}{{\theta^\alpha_l}}
\newcommand{\Expect}{{\mathbb{E}}}
\newcommand{\Tr}{{\rm Tr}}
\newcommand{\etc}{{\it etc~}}
\newcommand{\etal}{{\it etal~}}
\newcommand{\xset}{\mathbf{X}}
\newcommand{\fl}{\texttt{fl}}
\newcommand{\pdata}{\mathbf{\pi}}
\newcommand{\q}{\mathbf{q}}
\newcommand{\epdata}{\mathbf{\hat{\pi}}}
\newcommand{\gammaset}{\boldsymbol{\Gamma}}
\newcommand{\ei}{{\mathbf{e}_l^\alpha}}
\newcommand{\vtheta}{{\boldsymbol{\theta}}}
\newcommand{\sigmag}{{\nu}}
\newcommand{\sigmai}[2]{{\sigma^{#2}_{#1}}}
\newcommand{\qi}[1]{{q^{\alpha_{#1}}_{#1}}}
\newcommand{\BAS}{Bars-and-Stripes}
\newcommand{\circled}[1]{\raisebox{.5pt}{\textcircled{\raisebox{-.9pt} {#1}}}}
\newcommand{\qexpect}[1]{{\left\langle #1\right\rangle}}
\newcommand{\expect}[2]{{\mathop{\mathbb{E}}\limits_{\substack{#2}}\left[#1\right]}}
\newcommand{\var}[2]{{\mathop{\mathrm{Var}}\limits_{\substack{#2}}\left(#1\right)}}
\newcommand{\pshift}[1]{{p_{\thetav+#1}}}
\newcommand{\upcite}[1]{\textsuperscript{\cite{#1}}}
\newcommand{\Eq}[1]{Eq.~(\ref{#1})}
\newcommand{\Fig}[1]{Fig.~\ref{#1}}
\newcommand{\Lst}[1]{Listing.~\ref{#1}}
\newcommand{\Tbl}[1]{Table~\ref{#1}}
\newcommand{\Sec}[1]{Sec.~\ref{#1}}
\newcommand{\App}[1]{Appendix~\ref{#1}}
\newcommand{\bra}[1]{\mbox{$\left\langle #1 \right|$}}
\newcommand{\ket}[1]{\mbox{$\left| #1 \right\rangle$}}
\newcommand{\braket}[2]{\mbox{$\left\langle #1 | #2 \right\rangle$}}
\newcommand{\tr}[1]{\mathrm{tr}\mbox{$\left[ #1\right]$}}

\newcommand{\ra}[1]{\renewcommand{\arraystretch}{#1}}

%%%%%% Comment related
\newcommand{\red}[1]{[{\bf  \color{red}{ST: #1}}]}
\newcommand{\xred}[1]{[{\bf  \color{red}{\sout{ST: #1}}}]}
\newcommand{\green}[1]{[{\bf  \color{green}{XG: #1}}]}
\newcommand{\xgreen}[1]{[{\bf  \color{green}{\sout{XG: #1}}}]}
\newcommand{\blue}[1]{[{\bf  \color{blue}{JG: #1}}]}
\newcommand{\xblue}[1]{[{\bf  \color{blue}{\sout{JG: #1}}}]}
\newcommand{\cyan}[1]{[{\bf  \color{cyan}{ML: #1}}]}
\newcommand{\xcyan}[1]{[{\bf  \color{cyan}{\sout{ML: #1}}}]}
\newcommand{\purple}[1]{[{\bf  \color{purple}{MC: #1}}]}
\newcommand{\xpurple}[1]{[{\bf  \color{purple}{\sout{MC: #1}}}]}
\newcommand{\material}[1]{\iffalse[{\bf  \color{cyan}{Material: #1}}]\fi}

\newcounter{example}
\newenvironment{example}[1][]{\refstepcounter{example}\par\medskip
   \noindent \textbf{Example~\theexample. #1} \rmfamily}{\medskip}

%\newtheorem{theorem}{\textit{Theorem}}
%\newtheorem{corollary}{\textit Branching Rule}
%\theoremstyle{definition}\newtheorem{definition}{\textit{Definition}}
%\newtheorem{defin}[thm]{Definition}

\makeatother

%\externaldocument{ex_supplement}

\title{Computing solution space properties of combinatorial optimization problems via generic tensor networks
%\thanks{\funding{...}}
}

\author{
Jin-Guo Liu\thanks{Department of Physics, Harvard University, Cambridge, Massachusetts 02138, USA; QuEra Computing Inc., 1284 Soldiers Field Road, Boston, MA, 02135, USA  %\thanks{QuEra Computing Inc., 1284 Soldiers Field Road, Boston, MA, 02135, USA}
    (\email{jinguoliu@g.harvard.edu}).}
\and {Xun Gao}\thanks{Department of Physics, Harvard University, Cambridge, Massachusetts 02138, USA (\email{xungao@g.harvard.edu}, contributed equally with Jin-Guo Liu to this work.).}
\and Madelyn Cain\footnotemark[2]
\and Mikhail D. Lukin\footnotemark[2]
\and Sheng-Tao Wang\thanks{QuEra Computing Inc., 1284 Soldiers Field Road, Boston, MA, 02135, USA}
}

@@

To this end, we show how to obtain all of these seemingly unrelated properties in a unified approach using
\textit{generic tensor networks}. Tensor networks are a computational model widely used in condensed matter physics~\cite{Orus2014}, quantum computing~\cite{Markov2008}, big data~\cite{Cichocki2014} and mathematics~\cite{Oseledets2011}.
They are also known as the sum-product networks in probabilistic modeling~\cite{Bishop2006} or \texttt{einsum} in linear algebra libraries such as NumPy~\cite{Harris2020}.
%\purple{I’d write out $2^{{\rm tw}(G)}$}\blue{I feel it is hard to explain line graph without texts.}
Recent progress in simulating quantum circuits with tensor networks~\cite{Gray2021, Pan2021, Kalachev2021} makes it possible to contract a randomly structured sparse tensor network with up to thousands of tensors in a reasonable time.
%\purple{Is the scaling of these algorithms known?}\blue{They are not exact algorithms, the time can be any, typically a few minutes, will it be helpful to mention this?}
In previous studies, the data types of the tensor elements are typically restricted to standard number types such as real numbers and complex numbers.
Here, we extend to \textit{generic tensor networks} by generalizing the tensor element data types to any type that has the algebraic structure of a commutative semiring.
%These two algorithms are related but are not equivalent.
%Dynamic programming is more general purpose, while a tensor network has a richer algebraic structure, and this structure is crucial for solution space property computation.
%\purple{Not clear how first part and second part of sentence are related}\blue{now? or maybe just removing the comparison with dynamic programming?}
%For sparse graphs, the treewidth is usually much smaller than the number of vertices~\cite{Fomin2006}. Our algorithms, on the other hand, are much more versatile than traditional methods and can be used to compute many other properties than just the MIS size.
%The independent set problem is related to physics applications such as in the hard-core lattice gas model~\cite{Dyre2016, Fernandes2007}
%in statistical mechanics and the Rydberg hamiltonian with neutral atoms~\cite{Pichler2018, Ebadi2022};
%Property computation can, for example, be used to understand phase transitions~\cite{Dyre2016, Fernandes2007},
%to identify harder graphs in an ensemble of graphs~\cite{Ebadi2022} and to analyze the presence of the overlap gap property~\cite{Gamarnik2013, Gamarnik2019} related to the clustering phenomenon of large independent sets.
%However, there is a lack of general and versatile tool to compute these different properties of independent sets.
%To keep the discussion focused, 
In what follows, for clarity of presentation, we focus on the independent set problem in the main text, and show how to compute the solution space properties for other combinatorial optimization problems in  \App{sec:otherproblems}. The latter include cutting, matching, vertex coloring, satisfiability,  dominating set,  set packing,  set covering, and the clique problem.
%\xpurple{More details on which combinatorial optimization problems this applies to?}

The paper is organized as follows.
We first introduce the basic concepts of tensor networks and generic programming in \Sec{sec:tn} and \Sec{sec:generic}.
Then we show how to reduce the independent set problem to a tensor network contraction problem in \Sec{sec:tnmap}. 
Subsequently, we explain how to engineer the element types to compute various solution space properties in \Sec{sec:counting}, \Sec{sec:enumeration}, and \Sec{sec:weighted}. % to count independent sets at a given size, enumerate/sample independent sets at a given size and find largest set sizes in weighted graphs, respectively.
% by computing a number of independent set properties of certain sparse graphs on central processing units (CPUs) and graphics processing units (GPUs) and show its good performance.
Lastly, we provide three example applications in \Sec{sec:examples} to demonstrate the versatility of our tool.
A benchmark to demonstrate the performance of our algorithms can be found in both the supplementary material and the code repository.
%In the first example, we show how our method can help computing the entropy constant for some hardcore lattice gases on 2D square lattices. We compute the 2D generalization of Fibonacci integer sequence up to size $39$, which is larger than the previous record $37$.
%In the second example, we analyze the pairwise Hamming distances from configurations directly sampled from large independent sets of King's graphs at $0.8$ filling and three regular graphs. We show the strongest numeric evidence so far of the absence of overlap gap property in King's graph and the existence of Overlap Gap properties in three regular graphs.
%In the last example, we show how our algorithm helps people to understand the performance of a quantum variational algorithms on Rydberg atom arrays to find maximum independent sets. This is a research frontier that understanding which can help people design better quantum algorithms to beat its classical counterpart.
%Our method can also be used to find ``maximal'' independent set properties; 
%a maximal independent set is an independent set that is not a subset of any other independent set, but its size may not be the maximum.

\section{Tensor networks}\label{sec:tn}
A tensor network~\cite{Cirac2021, Orus2014} is composed of a collection of tensors, a collection of labels associated with tensor dimensions, and an indicator of which tensor is the output.
The operation to evaluate a tensor network is called \textit{contraction}, which is defined as a summation of tensor element products over the labels not appearing in the output tensor.
For example, the matrix multiplication is a special tensor network that can be represented as $C_{ik} = A_{ij}B_{jk}$, where $A, B$ and $C$ are matrices (two-dimensional tensors), $ik,ij$ and $ik$ are labels associated with them, and ``$=$'' is the indicator of $C$ being the output, while its contraction is defined as $C_{ik} = \sum_j A_{ij}B_{jk}$.
%can be viewed as a generalization of binary matrix multiplication to nary tensor contraction.
%Generalized Einstein's notation is often used to represent a tensor network, where we use a label in the subscript to represent a degree of freedom.
%One can enumerate these degrees of freedom and accumulate the product of tensor elements to the output tensor.
% Einstein's notation is often used to represent a tensor network, e.g.\ it represents the matrix multiplication between two matrices $A$ and $B$ as $C_{ik} = A_{ij}B_{jk}$,
% where we use a label in the subscript to represent a degree of freedom.
% One can enumerate these degrees of freedom and accumulate the product of tensor elements to the output tensor.
% In the standard Einstein's notation for tensor networks in physics, each index appears precisely twice, either both in input tensors (which will be summed over) or one in a input tensor and another in the output tensor.
% Hence a tensor network can be represented as a open simple graph,
% where a tensor is mapped to a vertex, a label shared by two input tensors is mapped to a normal edge and a label appears in the output tensor is mapped to an open edge.
% In this work, we adopt a more general form of tensor networks that do not restrict how many times a label can appear in tensors. 
The graphical representation of a tensor network is an open (meaning an edge can connect to external vertices) hypergraph, where an input tensor is mapped to a vertex and a label is mapped to a hyperedge that can connect an arbitrary number of vertices, while the labels appearing in the output tensor are open.
Our notation is a minor generalization of the standard tensor network notation used in physics as we do not restrict the number of times a label can appear in the tensors. 
While this generalized form is equivalent in representation power, it can have smaller contraction complexity as will be illustrated in the supplementary material.
% because one can easily translate a generalized tensor network to the standard notation by adding $\delta$ tensors, a high dimensional equivalence of the identity matrix.
%However, introducing $\delta$ tensors may significantly increase the complexity of the tensor contraction. We illustrate this subtle point in \App{sec:tensorbad}.

@@

This motivates us to identify the most general tensor element type allowed in a tensor network contraction program.
We find that as long as the tensor elements are members of a commutative semiring, the tensor network contraction will be well defined and the result will be independent of the contraction order.
Here, a commutative semiring is a field that needs not to have an additive inverse and a multiplicative inverse.
Giving up these nice properties of fields has significant implications for tensor computation: tensor network compression algorithms might not be applicable because matrix factorization is NP-hard for commutative semirings~\cite{Shitov2014} and matrix multiplication faster than $O(n^3)$ does not exist for an algebra without an additive inverse~\cite{Kerr1970}.
Here, we only use the commutative properties of an algebra for optimizing the tensor network contraction order.
To define a commutative semiring with the addition operation $\oplus$ and the multiplication operation $\odot$ on a set $S$, the following relations must hold for any arbitrary three elements $a, b, c \in S$.
\begin{align*}
(a \oplus b) \oplus c = a \oplus (b \oplus c) & \hspace{5em}\text{$\triangleright$ commutative monoid $\oplus$ with identity $\mymathbb{0}$}\\
a \oplus \mymathbb{0} = \mymathbb{0} \oplus a = a &\\
a \oplus b = b \oplus a &\\
&\\
(a \odot b) \odot c = a \odot (b \odot c)  &   \hspace{5em}\text{$\triangleright$ commutative monoid $\odot$ with identity $\mymathbb{1}$}\\
a \odot  \mymathbb{1} =  \mymathbb{1} \odot a = a &\\
a \odot b = b \odot a &\\
&\\
a \odot (b\oplus c) = a\odot b \oplus a\odot c  &  \hspace{5em}\text{$\triangleright$ left and right distributive}\\
(a\oplus b) \odot c = a\odot c \oplus b\odot c &\\
&\\
a \odot \mymathbb{0} = \mymathbb{0} \odot a = \mymathbb{0}
\end{align*}
%We require the algebra being multiplicative commutative because the contraction order optimizer will change the multiplication order of elements, while we require that the contraction order does not change the contraction results.
In the following sections, we will show how to compute solution space properties of independent sets using the same tensor network contraction algorithm by engineering tensor element algebra.
The Venn diagram in \Fig{fig:venn-diagram} shows the different types of algebra we will introduce in the main text and their relation, and Table~\ref{tbl:generictypes} summarizes which solution space properties can be computed by which tensor element types.

@@

    \caption{The tensor network element types used in this work and their relations.
    The overlap between two ellipses indicates that a new algebra can be created by combining those two types of algebra. ``Largest order'' and ``Largest 2 orders'' mean truncating the polynomial by only keeping its largest or largest two orders.
    The purposes of these element types can be found in \Tbl{tbl:generictypes}.}
     \label{fig:venn-diagram}
\end{figure}

\begin{table}[t!]\centering
\begin{minipage}{\columnwidth}
\ra{1.3}
        \begin{tabularx}{\textwidth}{sb}\toprule
            \hline
   \textbf{Element type}     & \textbf{Solution space property} \\
   {$\mathbbm{R}$}     & {Counting of all independent sets} \\
   {Polynomial} (\Eq{eq:polynomial})     & {Independence polynomial} \\
   {Tropical (\Eq{eq:tropical})}    & {Maximum independent set size} \\
   {Extended tropical of order $k$ (\Eq{eq:ext-tropical})}    & {Largest $k$ independent set sizes} \\
   {Polynomial truncated to $k$-th order (\Eq{eq:countingtropical} and \Eq{eq:max2poly})}     & {$k$ largest independent sizes and their degeneracy} \\
   {Set} (\Eq{eq:set})     & {Enumeration of independent sets} \\
   {Sum-Product expression tree} (\Eq{eq:expr})     & {Sampling of independent sets} \\
   {Polynomial truncated to largest order combined with bit string} (\Eq{eq:singleconfig})     & {Maximum independent set size and one of such configurations} \\
   {Polynomial truncated to $k$-th order combined with set} (\Eq{eq:countingtropicalset})    & {$k$ largest independent set sizes and their enumeration} \\
            \bottomrule
        \end{tabularx}
    \caption{Tensor element types and the independent set properties that can be computed using them.}\label{tbl:generictypes}
\end{minipage}
\end{table}

>>>>

\documentclass[onefignum, onetabnum]{siamart190516}

\usepackage[linesnumbered, ruled, vlined, algo2e]{algorithm2e}
\input{shared}

\title{Computing solution space properties of combinatorial optimization problems via generic tensor networks}

\externaldocument{suppl}

@@

To this end, we show how to obtain all of these seemingly unrelated properties in a unified approach using
\textit{generic tensor networks}. Tensor networks are a computational model widely used in condensed matter physics~\cite{Orus2014}, quantum computing~\cite{Markov2008}, big data~\cite{Cichocki2014} and mathematics~\cite{Oseledets2011}.
They are also known as the sum-product networks in probabilistic modeling~\cite{Bishop2006} or \texttt{einsum} in linear algebra libraries such as NumPy~\cite{Harris2020}.
%\purple{I’d write out $2^{{\rm tw}(G)}$}\blue{I feel it is hard to explain line graph without texts.}
Recent progress in simulating quantum circuits with tensor networks~\cite{Gray2021, Pan2021, Kalachev2021} makes it possible to contract a randomly structured sparse tensor network with up to thousands of tensors in a reasonable time.
%\purple{Is the scaling of these algorithms known?}\blue{They are not exact algorithms, the time can be any, typically a few minutes, will it be helpful to mention this?}
In previous studies, the data types of the tensor elements are typically restricted to standard number types such as real numbers and complex numbers.
Here, we extend to \textit{generic tensor networks} by generalizing the tensor element data types to any type that has the algebraic structure of a commutative semiring.
%These two algorithms are related but are not equivalent.
%Dynamic programming is more general purpose, while a tensor network has a richer algebraic structure, and this structure is crucial for solution space property computation.
%\purple{Not clear how first part and second part of sentence are related}\blue{now? or maybe just removing the comparison with dynamic programming?}
%For sparse graphs, the treewidth is usually much smaller than the number of vertices~\cite{Fomin2006}. Our algorithms, on the other hand, are much more versatile than traditional methods and can be used to compute many other properties than just the MIS size.
%The independent set problem is related to physics applications such as in the hard-core lattice gas model~\cite{Dyre2016, Fernandes2007}
%in statistical mechanics and the Rydberg hamiltonian with neutral atoms~\cite{Pichler2018, Ebadi2022};
%Property computation can, for example, be used to understand phase transitions~\cite{Dyre2016, Fernandes2007},
%to identify harder graphs in an ensemble of graphs~\cite{Ebadi2022} and to analyze the presence of the overlap gap property~\cite{Gamarnik2013, Gamarnik2019} related to the clustering phenomenon of large independent sets.
%However, there is a lack of general and versatile tool to compute these different properties of independent sets.
%To keep the discussion focused, 
In what follows, for clarity of presentation, we focus on the independent set problem in the main text, and show how to compute the solution space properties for other combinatorial optimization problems in \Cref{sec:otherproblems}. The latter include cutting, matching, vertex coloring, satisfiability,  dominating set,  set packing,  set covering, and the clique problem.
%\xpurple{More details on which combinatorial optimization problems this applies to?}

The paper is organized as follows.
We first introduce the basic concepts of tensor networks and generic programming in \Cref{sec:tn} and \Cref{sec:generic}.
Then we show how to reduce the independent set problem to a tensor network contraction problem in \Cref{sec:tnmap}. 
Subsequently, we explain how to engineer the element types to compute various solution space properties in \Cref{sec:counting}, \Cref{sec:enumeration}, and \Cref{sec:weighted}. % to count independent sets at a given size, enumerate/sample independent sets at a given size and find largest set sizes in weighted graphs, respectively.
% by computing a number of independent set properties of certain sparse graphs on central processing units (CPUs) and graphics processing units (GPUs) and show its good performance.
Lastly, we provide three example applications in \Cref{sec:examples} to demonstrate the versatility of our tool.
A benchmark to demonstrate the performance of our algorithms can be found in both the \Cref{sec:benchmark} and the code repository.
%In the first example, we show how our method can help computing the entropy constant for some hardcore lattice gases on 2D square lattices. We compute the 2D generalization of Fibonacci integer sequence up to size $39$, which is larger than the previous record $37$.
%In the second example, we analyze the pairwise Hamming distances from configurations directly sampled from large independent sets of King's graphs at $0.8$ filling and three regular graphs. We show the strongest numeric evidence so far of the absence of overlap gap property in King's graph and the existence of Overlap Gap properties in three regular graphs.
%In the last example, we show how our algorithm helps people to understand the performance of a quantum variational algorithms on Rydberg atom arrays to find maximum independent sets. This is a research frontier that understanding which can help people design better quantum algorithms to beat its classical counterpart.
%Our method can also be used to find ``maximal'' independent set properties; 
%a maximal independent set is an independent set that is not a subset of any other independent set, but its size may not be the maximum.

\section{Tensor networks}\label{sec:tn}
A tensor network~\cite{Cirac2021, Orus2014} is composed of a collection of tensors, a collection of labels associated with tensor dimensions, and an indicator of which tensor is the output.
The operation to evaluate a tensor network is called \textit{contraction}, which is defined as a summation of tensor element products over the labels not appearing in the output tensor.
For example, the matrix multiplication is a special tensor network that can be represented as $C_{ik} = A_{ij}B_{jk}$, where $A, B$ and $C$ are matrices (two-dimensional tensors), $ik,ij$ and $ik$ are labels associated with them, and ``$=$'' is the indicator of $C$ being the output, while its contraction is defined as $C_{ik} = \sum_j A_{ij}B_{jk}$.
%can be viewed as a generalization of binary matrix multiplication to nary tensor contraction.
%Generalized Einstein's notation is often used to represent a tensor network, where we use a label in the subscript to represent a degree of freedom.
%One can enumerate these degrees of freedom and accumulate the product of tensor elements to the output tensor.
% Einstein's notation is often used to represent a tensor network, e.g.\ it represents the matrix multiplication between two matrices $A$ and $B$ as $C_{ik} = A_{ij}B_{jk}$,
% where we use a label in the subscript to represent a degree of freedom.
% One can enumerate these degrees of freedom and accumulate the product of tensor elements to the output tensor.
% In the standard Einstein's notation for tensor networks in physics, each index appears precisely twice, either both in input tensors (which will be summed over) or one in a input tensor and another in the output tensor.
% Hence a tensor network can be represented as a open simple graph,
% where a tensor is mapped to a vertex, a label shared by two input tensors is mapped to a normal edge and a label appears in the output tensor is mapped to an open edge.
% In this work, we adopt a more general form of tensor networks that do not restrict how many times a label can appear in tensors. 
The graphical representation of a tensor network is an open (meaning an edge can connect to external vertices) hypergraph, where an input tensor is mapped to a vertex and a label is mapped to a hyperedge that can connect an arbitrary number of vertices, while the labels appearing in the output tensor are open.
Our notation is a minor generalization of the standard tensor network notation used in physics as we do not restrict the number of times a label can appear in the tensors. 
While this generalized form is equivalent in representation power, it can have smaller contraction complexity as will be illustrated in \Cref{sec:tensorbad}.
% because one can easily translate a generalized tensor network to the standard notation by adding $\delta$ tensors, a high dimensional equivalence of the identity matrix.
%However, introducing $\delta$ tensors may significantly increase the complexity of the tensor contraction. We illustrate this subtle point in \Cref{sec:tensorbad}.

@@

This motivates us to identify the most general tensor element type allowed in a tensor network contraction program.
We find that as long as the tensor elements are members of a commutative semiring, the tensor network contraction will be well defined and the result will be independent of the contraction order.
Here, a commutative semiring is a field that needs not to have an additive inverse and a multiplicative inverse.
Giving up these nice properties of fields has significant implications for tensor computation: tensor network compression algorithms might not be applicable because matrix factorization is NP-hard for commutative semirings~\cite{Shitov2014} and matrix multiplication faster than $O(n^3)$ does not exist for an algebra without an additive inverse~\cite{Kerr1970}.
Here, we only use the commutative properties of an algebra for optimizing the tensor network contraction order.
To define a commutative semiring with the addition operation $\oplus$ and the multiplication operation $\odot$ on a set $S$, the following relations must hold for any arbitrary three elements $a, b, c \in S$.
\begin{align*}
(a \oplus b) \oplus c = a \oplus (b \oplus c) & \hspace{5em}\text{$\triangleright$ commutative monoid $\oplus$ with identity $\mymathbb{0}$}\\
a \oplus \mymathbb{0} = \mymathbb{0} \oplus a = a &\\
a \oplus b = b \oplus a &\\
&\\
(a \odot b) \odot c = a \odot (b \odot c)  &   \hspace{5em}\text{$\triangleright$ commutative monoid $\odot$ with identity $\mymathbb{1}$}\\
a \odot  \mymathbb{1} =  \mymathbb{1} \odot a = a &\\
a \odot b = b \odot a &\\
&\\
a \odot (b\oplus c) = a\odot b \oplus a\odot c  &  \hspace{5em}\text{$\triangleright$ left and right distributive}\\
(a\oplus b) \odot c = a\odot c \oplus b\odot c &\\
&\\
a \odot \mymathbb{0} = \mymathbb{0} \odot a = \mymathbb{0}
\end{align*}
%We require the algebra being multiplicative commutative because the contraction order optimizer will change the multiplication order of elements, while we require that the contraction order does not change the contraction results.
In the following sections, we will show how to compute solution space properties of independent sets using the same tensor network contraction algorithm by engineering tensor element algebra.
The Venn diagram in \Cref{fig:venn-diagram} shows the different types of algebra we will introduce in the main text and their relation, and \Cref{tbl:generictypes} summarizes which solution space properties can be computed by which tensor element types.

@@

    \caption{The tensor network element types used in this work and their relations.
    The overlap between two ellipses indicates that a new algebra can be created by combining those two types of algebra. ``Largest order'' and ``Largest 2 orders'' mean truncating the polynomial by only keeping its largest or largest two orders.
    The purposes of these element types can be found in \Cref{tbl:generictypes}.}
     \label{fig:venn-diagram}
\end{figure}

\begin{table}[t!]\centering
\begin{minipage}{\columnwidth}
\ra{1.3}
        \begin{tabularx}{\textwidth}{sb}\toprule
            \hline
   \textbf{Element type}     & \textbf{Solution space property} \\
   {$\mathbbm{R}$}     & {Counting of all independent sets} \\
   {Polynomial} (\Cref{eq:polynomial})     & {Independence polynomial} \\
   {Tropical (\Cref{eq:tropical})}    & {Maximum independent set size} \\
   {Extended tropical of order $k$ (\Cref{eq:ext-tropical})}    & {Largest $k$ independent set sizes} \\
   {Polynomial truncated to $k$-th order (\Cref{eq:countingtropical} and \Cref{eq:max2poly})}     & {$k$ largest independent sizes and their degeneracy} \\
   {Set} (\Cref{eq:set})     & {Enumeration of independent sets} \\
   {Sum-Product expression tree} (\Cref{eq:expr})     & {Sampling of independent sets} \\
   {Polynomial truncated to largest order combined with bit string} (\Cref{eq:singleconfig})     & {Maximum independent set size and one of such configurations} \\
   {Polynomial truncated to $k$-th order combined with set} (\Cref{eq:countingtropicalset})    & {$k$ largest independent set sizes and their enumeration} \\
            \bottomrule
        \end{tabularx}
    \caption{Tensor element types and the independent set properties that can be computed using them.}\label{tbl:generictypes}
\end{minipage}
\end{table}

\section{Tensor network representation of independent sets} \label{sec:tnmap}
This section describes the reduction of the independent set problem to a tensor network contraction problem.
An alternative interpretation, perhaps more accessible to physicists, can be found in \Cref{sec:energymodel}, where we introduce the reduction from the energy model of hardcore lattice gases~\cite{Dyre2016, Fernandes2007}.
Let $G=(V,E)$ be an undirected graph with each vertex $v\in V$ being associated with a weight $w_v$.
An independent set $I \subseteq V$ is a set of vertices that for any vertex pair $u,v \in I$, we have $(u, v) \not\in E$.
To reduce the independent set problem on $G$ to a tensor network contraction, we first map a vertex $v\in V$ to a label $s_v \in \{0, 1\}$ of dimension $2$, where we use $0$ ($1$) to denote a vertex is absent (present) in the set.
For each vertex $v$, we define a parameterized rank-one tensor indexed by $s_v$ as
\begin{equation}
    W(x_v^{w_v}) = \left(\begin{matrix}
        1 \\
        x_v^{w_v}
    \end{matrix}\right),\label{eq:vertextensor}
\end{equation}
where $x_v$ is a variable associated with $v$.
Tensor elements can be indexed by subscripts, e.g.\ $W(x_v^{w_v})_0=1$ is the first element associated with $s_v=0$ and $W(x_v^{w_v})_1=x_v^{w_v}$ is the second element associated with $s_v=1$.
Similarly, for each edge $(u, v) \in E$, we define a matrix $B$ indexed by $s_u$ and $s_v$ as
\begin{equation}
    \qquad \quad 
       B = \left(\begin{matrix}
        1  & 1\\
        1 & 0
    \end{matrix}\right), \label{eq:edgetensor}
\end{equation}
which encodes the independent set constraint since $B_{11} = 0$, meaning the two vertices cannot be both in the independent set.
